On the effects of partying on next-day sports performance

### 2017-06-13 writing/draft.tex
Moved figures around, added a figure caption, added more detailed analyses.

\subsection{Teams}
There has been an attempt to answer the question at the team-level before: i.e. do teams perform
worse after a night on the town, and if so does it depend on which city they 
visited.\citep{ezekowitz} However, the study only
looks at performance in party cities, as opposed to subsequent day performance (i.e. the study does not consider a delayed hangover effect). Moreover, their sample size was much smaller, because the author only considered Sunday afternoon games
from 2006-14. Finally, there was no methodology used behind the selection of party cities:
the authors first tried to define ad-hoc five ``party cities'', and when that did not yield
a significant result they tried estimating fixed-effects for each city. The author does conclude
with weak statistical evidence that teams systematically play worse on Sunday 
afternoons when on the road, regardless of the city.

@@

Our takeaway from the figure on page \ref{fig: next day opponent} is that indeed there is some correlation between last game location
and next day opponent, and in particular this is true for west coast teams. We hypothesize that
this is due to practical limitations on flight schedules. E.g. if a team is unwilling to take a red-eye
flight, then there is little hope to play an evening west coast game and make it to the east coast for a game the very next day;
the same is \emph{not true} when flying from east to west, in which case a morning flight is reasonable.
The fact that there is some correlation between last game location and next day opponent, even if only in west coast cities, increases the motivation for conditioning on spread, which accounts may help account for systematic biases to the extent that some of these next-day opponents are particularly (not) skilled.

@@

\paragraph{A discrete indicator}
Because famous athletes often enjoy a celebrity lifestyle, using conventional statistics
such as number of drinking establishments in each city does not work; athletes instead
tend to visit high end nightclubs. We again turned to player interviews for ideas,
and found two major insights. From our conversations, 
we obtained a listing of cities which were anecdotally
mentioned as having active night-life. Between players, the rank ordering of these
was consistent to the extent that Los Angeles and New York city were top. 
This motivates using a binary indicator for party when either of these two cities have been visited within 24-hours of another game.

\paragraph{A continuous measure}
We also found that athletes tend to party with other A-list celebrities,
particularly musicians. This motivated a continuous measure of nightlife, based on a simple count of the number of musicians in the county in which the stadium resides: if the last game was played at most 24-hours ago, use the log of the average number
of musicians in last game location, else emit zero.  
This captures
how much nightlife there is in a city since musical performances are correlated
with parties, if not an essential feature.

@@

\subsubsection{Placebo Tests} As a sanity check, we interact our
discrete-indicator for party city with an indicator for more than 24 hours of rest
between games, and call this our placebo effect. 
The idea is that with sufficient rest time, the hangover effect dissipates.

@@

\section{Analysis of Player Performance}
We additionally looked at actual performance metrics instead of relying on spread data. 
This allows us to avoid using a black-box feature, and also pin-point
more precisely what elements of the game effected by visiting a nightlife city. This motivates looking at number of points allowed, and number of points scored; we also considered how rebounds, fouls, and even injuries might be 
explained (in part) by a hangover effect.

@@

\begin{figure}   
  \centering
  \hspace{-55pt}
  \label{fig: next day opponent mlb}
  \includegraphics[scale=0.5]{next_day_opponent_mlb} 
\end{figure}

\begin{sidewaysfigure}   
  \centering
  \label{fig: why drinks or pop dont work}
  \includegraphics[scale=0.5]{why_drinks_or_population_dont_work} 
  \caption{We remark why using a statistic such as number of drinking establishments or 
    population size doesn't work. Although both of these are linearly correlated with musicians, 
    and even in spite of the extremal observations being ranked consistently across each statistic, 
    i.e. Charlotte Bobcats at the bottom and Los Angeles at the top,  
    the problem is that population and number of drinking establishments have sinusoidal 
    relationships with respect to the rank ordering imposed by the musicians index 
    that do not properly reflect variation in nightlife between cities. For example, both 
    incorrectly rank Detroit Michigan as having a high nightlife index. Another example, population incorrectly places New Orleans, Louisiana as having a low nightlife index, when in reality it is famous for its vibrant city life.}
\end{sidewaysfigure}

>>>>

\subsection{Teams}
There has been an attempt to answer the question at the team-level before: i.e. do teams perform
worse after a night on the town, and if so does it depend on which city they 
visited.\citep{ezekowitz} However, the study only
looks at performance in party cities, as opposed to subsequent day performance (i.e. the study does not consider a delayed hangover effect). Moreover, their sample size was much smaller, because the author only considered Sunday afternoon games
from 2006-14. Finally, there was no methodology used behind the selection of party cities:
the authors first tried to define ad-hoc five ``party cities'', and when that did not yield
a significant result they tried estimating fixed-effects for each city. The author does carry out a binomial test of how many home-game covers vs. away-game covers were met, 
and concludes
with weak statistical evidence that teams systematically play worse on Sunday afternoons 
when on the road, regardless of city.

@@

We plot a heatmap which displays the logged count of co-occurences between last game
location and current game location, split by home-and-away games.
Our takeaway from this figure is that indeed there is some correlation between last game location
and next day opponent (more so when both are away games,  
and in particular this is true for west coast teams). We hypothesize that
this is due to practical limitations on flight schedules. E.g. if a team is unwilling to take a red-eye
flight, then there is little hope to play an evening west coast game and make it to the east coast for a game the very next day;
the same is \emph{not true} when flying from east to west, in which case a morning flight is reasonable.
The fact that there is some correlation between last game location and next day opponent, even if only in west coast cities, increases the motivation for conditioning on spread, which may help account for systematic biases to the extent that some of these next-day opponents are particularly (not) skilled.

@@

\paragraph{A discrete indicator}
Because famous athletes often enjoy a celebrity lifestyle, using conventional statistics
such as number of drinking establishments in each city or population does not work (see appendix \ref{fig: why drinks or pop dont work}), athletes instead
tend to visit high end nightclubs. We again turned to player interviews for ideas,
and found two major insights. From our conversations, 
we obtained a listing of cities which were anecdotally
mentioned as having active night-life. Between players, the rank ordering of these
was consistent to the extent that Los Angeles and New York city were top. 
This motivates using a binary indicator for party when either of these two cities have been visited within 24-hours of another game.

\paragraph{A continuous measure}
We also found that athletes tend to party with other A-list celebrities,
particularly musicians. This motivated a continuous measure of nightlife, based on a simple count of the number of musicians in the county in which the stadium resides: if the last game was played at most 24-hours ago, use the log of the average number
of musical groups, sound recording studios, and musical publishers in county of the 
last game location, else emit zero.  
This captures
how much nightlife there is in a city since musical performances are correlated
with parties, if not an essential feature.

@@

\subsubsection{Placebo Tests} As a sanity check with our NBA data, we interact a
discrete-indicator for Los Angeles and New York with an 
indicator for \emph{more} than 24 hours of rest
between games, and call this our placebo effect. The only difference between this feature
and our discrete indicator for nightlife is the time-window of interest.
The idea is that with sufficient rest time, the hangover effect dissipates.

@@

\section{Analysis of Player Performance}
Within the NBA, we additionally looked at actual performance metrics instead of relying on point-spread data. 
This allows us to avoid using a black-box feature, and also pin-point
more precisely what elements of the game are affected by visiting a nightlife city. This motivates looking at number of points allowed, and number of points scored; we also considered how rebounds, fouls, and even injuries might be 
explained (in part) by a hangover effect.

@@

\begin{figure}   
  \centering
  \hspace{-55pt}
  \label{fig: next day opponent mlb}
  \includegraphics[scale=0.5]{next_day_opponent_mlb} 
  \caption{For the MLB, there are some differences when compared with the analogous
  plot for NBA. In particular, notice that a common scenario is for a team to play two away games 
  back to back at the same location. This is an artifact of baseball match-ups being played 
  in series as opposed to one-off games like in the NBA. We also notice in back-to-back
  away games that the sparse rows 
  (i.e. locations for which next-day opponent is restricted to only a strict subset of the 
  29 available teams to play) all are attributed by West coast teams. In particular, 
  Washington Nationals, San Francisco Giants, Oakland Athletics, Los Angeles Dodgers, and 
  Arizona Diamondbacks are locations from which a next-day game on the East coast are not possible.}
\end{figure}

\begin{sidewaysfigure}   
  \centering
  \label{fig: why drinks or pop dont work}
  \includegraphics[scale=0.5]{why_drinks_or_population_dont_work} 
  \caption{We remark why using a statistic such as number of drinking establishments or 
    population size doesn't work. Although both of these are linearly correlated with musicians, 
    and even in spite of the extremal observations being ranked consistently across each statistic, 
    i.e. Charlotte Bobcats toward the bottom and Los Angeles more or less at the top,  
    the problem is that population and number of drinking establishments have sinusoidal 
    relationships with respect to the rank ordering imposed by the musicians index 
    that do not properly reflect variation in nightlife between cities. For example, both 
    incorrectly rank Detroit Michigan as having a high nightlife index. Another example, 
    population incorrectly places New Orleans, Louisiana as having a low nightlife index, 
    when in reality it is famous for its vibrant city life.}
\end{sidewaysfigure}
